# Antecedents {#sec-why-antecedents}

Spatial econometrics may be seen in a number of different ways, as misguided,
as an inward-looking endeavour for the specially interested, but these views 
are rarely placed in their context. In this section, we will
sketch the antecedents of spatial econometrics, both providing views
on the research context when it came into being, and on changes in
econometrics over time. Grasping spatial econometrics today without
insight into its antecedents is unnecessarily difficult, so taking time
to explore how ideas came into being and developed is well-spent. 
  \index{spatial econometrics!history of}

In the same way that @fujitaetal:99 begin their study of the
spatial economy by looking at the antecedents of their subject, it
is helpful to place spatial econometrics in its temporal and academic
context. This context is sufficiently different from the contemporary
setting that it may be hard to grasp the background for many of the
features of spatial econometrics that came into being during its earlier
years. Indeed, the ranges of topics that were studied in economics in
the 1960's and 1970's differ markedly from those in focus today. If we
can sketch the context within which spatial econometrics was created,
and its methods developed, we should be able to illuminate choices
made then which influence our understanding and application of spatial
econometric methods.
  \index{spatial econometrics}
  \index{space economy}

Critics of the practice of spatial econometrics, such as
@gibbons+overman:12, appear to overlook these antecedents, and
consequently judge the potential of the field on a partial, perhaps
anachronistic, understanding, viewing phenomena with a history in
ahistorical way. Since we are attempting to provide an introduction to
applied spatial econometrics, we need to throw light on the original
motivations and concerns of the first scholars engaged in the field.
@anselin:10 indicates clearly and repeatedly [@anselin:88;
@anselin:06; @anselin:10a] that we should acknowledge *Spatial
Econometrics* by @paelinck+klaassen:79 of the Netherlands Economic
Institute as our starting point, and so celebrated thirty years of
spatial econometrics in 2009. This firm confirmation of the importance
of Jean Paelinck's contributions as scholar and community-builder
is fully justified. We should then turn to the motivations given in
@paelinck+klaassen:79 to indicate which contextual factors were
of importance at that time, and the breadth of the academic communities
with which they were in contact.
  \index{spatial econometrics!history of}

In a recent short commentary, @paelinck:13 [p. 11] recalled his
conviction, expressed in 1967, that 

> early econometric exercises ...
relating only variables posessing the same regional index ... were
inadequate to represent the correct spatial workings of the economy,
which would then be reflected in the policy outcomes. 

A year before,
@paelinck:12 [p. 36] pointed to salient isomorphisms linking
spatial regression models, simultaneous equation models and input-output
models; these were known of and discussed in the early formative period
of spatial econometrics. We will return in subsequent chapters to the
ways in which spatial regression models may be specified, but for now,
a simple presentation of these isomorphisms as perceived in the early
period is sufficient:
  \index{spatial econometrics!history of}

```{r, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) {
cat("\\begin{equation*}\n{\\mathbf y} = {\\mathbf A}{\\mathbf y} + {\\mathbf X}{\\mathbf b} + \\mbfvarepsilon\n\\end{equation*}\n")
} else {
cat("\\begin{equation*}\n{\\mathbf y} = {\\mathbf A}{\\mathbf y} + {\\mathbf X}{\\mathbf b} + {\\mathbf \\varepsilon}\\end{equation*}\n")
}
```
\noindent
is a spatial regression model where ${\mathbf A}$ is a matrix expressing
the mutual first order spatial dependencies between regions --- the
similarity of this form and the Koyck distributed lag model is striking
[@koyck:54; @klein:58; @griliches:67];
  \index{spatial regression model}
  \index{distributed lag model}

```{r, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) {
cat("\\begin{equation*}\n{\\mathbf A}{\\mathbf y} + {\\mathbf X}{\\mathbf b} = \\mbfvarepsilon\n\\end{equation*}\n")
} else {
cat("\\begin{equation*}\n{\\mathbf A}{\\mathbf y} + {\\mathbf X}{\\mathbf b} = {\\mathbf \\varepsilon}\n\\end{equation*}\n")
}
```
\noindent
is a simultaneous equation model where ${\mathbf A}$ is a matrix expressing the dependencies between the equations; and:
  \index{simultaneous equation model}

\begin{equation*}
{\mathbf y} = {\mathbf A}{\mathbf y} + {\mathbf f}
\end{equation*}
\noindent
is an input-output model where ${\mathbf A}$ is a matrix of sectoral 
input-output coefficients, and ${\mathbf f}$ is final demand.
  \index{input-output model}

Input-output models, simultaneous equation models, and the importance of
policy outcomes were all known intimately at the Netherlands Economic
Institute at this time, and elsewhere among applied economists. The
isomporhisms flowed from the known to the unknown, from the stuff of
contemporary research and policy advice to doubts about the calibration
of aspatial models, and on to what became termed spatial econometrics. If
we compare these topics with those described for Regional Science by
@boyce:04, we can see the outlines of research priorities at the
time: including urban and regional models for planning, regional and
interregional input-output models, transport and location models. During
the 1960s and 1970s, many of these models were enhanced, matching needs
for policy advice, to cover environmental questions, adding natural
resources as inputs and pollution to outputs. Paelinck's co-author in a
key paper in spatial econometrics [@hordijk+paelinck:76], went on
to work in environmental management and research.
  \index{Regional Science}
  \index{spatial econometrics!history of}

Reading @paelinck+klaassen:79 [p. viii], we see that the programme
of research into the space economy undertaken at the Netherlands Economic
Institute led first to the publication of @paelinck+nijkamp:75,
and then to @klaassenetal:79, published in the same year as
*Spatial Econometrics*. All three books were published in the
same series and appear to reflect the core concerns of economists at
the Institute doing research on regionalised national macro-economic
models. The direct link to Jan Tinbergen is evident in the account
of the context of economic research in the Netherlands given by
@theil:64 [p. 49]. If we take Paelinck at his word, he and
his colleagues were aware that an aspatial regionalisation of national
accounts, of input-output models, or transport models, might prejudice
policy advice and outcomes through inadequate and inappropriate
calibration.
  \index{Regional Science}
  \index{spatial econometrics!history of}

@klaassenetal:79 is mainly concerned with model construction, while
about a third of @paelinck+nijkamp:75 is devoted to input-output
analysis. Both books show sustained concern for aggregate economic measurement,
especially of national accounts data, intersectoral transactions, and many
other topics. Considerable attention is also paid to the data collection
units, be they sectors or regions.  The need to attempt to define regions
that match the underlying economic realities was recognised clearly, and
a key part of @paelinck+nijkamp:75 is devoted to regionalisation,
and the distinction between functional regions and homogenous regional
classifications is made. In the motivation for spatial econometric
models given in @paelinck+klaassen:79 [pp. 2--3], consumption
and investment in a region are modelled as depending on income both in
the region itself and in its contiguous neighbours, termed a "spatial
income-generating model." It became important to be able to calibrate
planning models of this kind to provide indications of the possible
outcomes of alternative policy choices, hence the need for spatial
econometrics.
  \index{Regional Science}
  \index{spatial econometrics!history of}
  \index{regionalisation}
  \index{functional region}

Economic planning was widespread in Europe at the time, and was also
central in the development of Regional Science, in particular input-output
models; as @boyce:04 recounts, Walter Isard worked closely with
Wassily Leontief. Operational and planning motivations for applied
economics were unquestioned, as economists in the post-war period saw
their role, beyond educating young economists, as providing rational
foundations for economic policy. It is worth noting that Jean Paelinck
participated actively in the Association de Science Régionale De Langue
Française, becoming president in 1973--1976. The first president of
the association was François Perroux, who had founded it with Walter
Isard in 1961 [@baillyetal:12].
  \index{input-output model}
  \index{Regional Science}

Until the 1980s, it was not at all unusual to publish original
results in other languages than English. French spatial economic
research, for example @ponsard:83, while making little impact
in Anglophone countries, was widely used in teaching and research
elsewhere [@billot+thisse:92]. They contrast, though, the "word
wizardry of François Perroux with the rigour of Claude Ponsard"
@billot+thisse:92 [p. 194], echoing the views expressed by
@dreze:64 [pp. 3--4] with regard to the work of Perroux. Even
if we accept that "word wizardry" deserves more rigour and recasting
in normative and empirically testable forms, it is also part of the
context within which spatial econometrics came into being. A reading of
@perroux:50 [a lecture given at Harvard University] is worthwhile,
because it not only gives the reader a vignette of the context in
the post-war period, but also provides a discussion of economic space,
as opposed to "banal, unreflected space, mere position", that has
largely disappeared from our considerations.
  \index{Regional Science}
  \index{space economy}

The title of the journal: *Regional and Urban Economics,
Operational Methods*, founded by Jean Paelinck in 1971, and which
was renamed as *Regional Science and Urban Economics* in 1975.
@boyce:04 [p. 43], points to the perceived importance of
"operational methods", a version of the term "operational theory
and method" used in the title of @paelinck+nijkamp:75. Spatial
econometrics does not seem to have come into being as a set of estimation
techniques as such, as perhaps we might think today, but rather as
an approach addressing open research questions both in space economy
and in the enhancement of interregional models to be used in offering
policy advice.
  \index{Regional Science}
  \index{spatial econometrics!history of}

Were motivations of this kind common during the 1960s and early
1970s? Not only was the spread of Regional Science extensive and
firmly established [@boyce:04], but public bodies were concerned
to regionalise economic measurement and policy advice [see,
for example @graham+romans:71]. In Britain, *Environment
and Planning* was started in 1969 with Alan Wilson as founding
editor and published by Pion; he was assistant director at the
Centre for Environmental Studies at this time before moving to
the University of Leeds. In a recently published lecture series,
@wilson:12 [pp. 35-36] cites @paelinck+nijkamp:75 as
giving principles for contributions from economics to urban and regional
analysis [see also @wilson:00, p. 34].  The papers presented at
annual Regional Science meetings were published in a series by Pion; the
first number in the series included contributions by @granger:69
and @cliff+ord:69.
  \index{Regional Science}
  \index{spatial econometrics!history of}
  \index{economic measurement}

In a contribution to a panel session at the 2006 annual meeting of
the American Association of Geographers (co-panelists Luc Anselin
and Daniel Griffith), Keith Ord pointed to the continued relevance
of Granger's remarks at the meeting almost fourty years earlier
[@ord:10, pp. 168-169]; we will return to these concerns below. As
noted by @bivand:08 [pp. 1--5], communities of researchers working
in and near mathematical and theoretical geography was more integrated
in the pre-internet and pre-photocopier age than one might expect, with
duplicated working papers prepared using stencils circulating rapidly
between collaborating academic centres. Knowledge of the preliminary
results of other researchers then fed through into rapid innovation
in an exciting climate for those with access to these meetings and
working papers.

There was considerable overlap between quantitative geography and
regional science, so that work like @cliff+ord:69 is cited by
@hordijk:74, and was certainly known at the Netherlands Economic
Institute several years earlier. Although it has not been possible
to find out who participated in the August 1968 conference of the
British and Irish Section of the Regional Science Association at which
@cliff+ord:69 was read, it was not unusual for members of other
sections to be present, and to return home with bundles of duplicated
papers. Up to the 1990s, presenters at conferences handed out copies of
their papers, and conference participants posted home parcels of these
hand-outs, indexed using the conference programme.
  \index{Regional Science}
  \index{quantitative geography}

Leslie Hepple was among the more thorough scholars working on the
underpinnings of spatial econometrics prior to the publication of
@paelinck+klaassen:79. His wide-ranging review @hepple:74
is cited by @bartels+hordijk:77, again demonstrating the close
links between those working in this field. We will be returning to
the review paper, and to @hepple:76, which studies methods of
estimation for spatial econometrics models in some depth, building on
and extending @ord:75.
  \index{Regional Science}
  \index{spatial econometrics!history of}
  \index{quantitative geography}

@hepple:74, like @cliff+ord:73, saw no distinction between
spatial statistics and the antecedents to spatial econometrics. Obviously,
spatial econometrics was strongly influenced by the research tasks
undertaken by regional and urban economists and regional scientists. As
@griffith+paelinck:07 [p. 211] point out, spatial statistics and
spatial econometrics continue to share most topics of interests, with each
also possessing shorter lists of topics that have been of less concern to
the other. They advocate a "non-standard" spatial econometrics, which
is inclusive to wider concerns. It seems appropriate in this context to
mention the somewhat heterodox position taken by @mcmillen:10,
who draws attention to the crucial issue of functional form, which he
argues may well lie behind observed spatial autocorrelation; we will
return to this in later chapters.
  \index{spatial econometrics!history of}
  \index{spatial statistics}


